---
title: "Whittaker biomes - the dataset"
author: "Valentin Stefan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Whittaker biomes - the dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**This document explains how the `Whittaker_biomes` data frame was constructed.**

# Table of contents
- [Original graph](#Original_graph)
- [Processing PDF](#PDF)
- [Importing PostScript file in R with `grImport`](#grImport)
    - [Read in the PostScript file](#ReadPS)
- [Visualize and clean (subset) the "Picture" object](#picture)
- [Get the PostScript coordinates from the cleaned "Picture" object](#PScoordinates)
- [Convert PostScript coordinates into meaningful coordinates](#conversion_1)
    - [Convert between coordinates systems](#conversion_2)
- [Prepare Whittaker_biomes data frame](#prepare)
    - [Get the colors also](#colors)
- [Examples](#examples)

<div id='Original_graph'/>
## Original graph

The original graph is Figure 5.5 in *Ricklefs, R. E. (2008), The economy of nature. W. H. Freeman and Company.* (Chapter 5, Biological Communities, The biome concept).

<div id='PDF'/>
## Processing PDF

A PDF page containing the Whittaker biomes graph was imported in [Inkscape](https://inkscape.org/en/). Text and extra graphic layers where removed and the remaining layers were exported as PostScipt file format (`File > Save As > Save as type: > PostScript (*.ps)`). Note that a multi-page PDF document can be split into component pages with [PDFTK Builder](https://www.pdflabs.com/tools/pdftk-the-pdf-toolkit/) application. 

<div id='grImport'/>
## Importing PostScript file in R with `grImport`

The main steps of importing PostsScipt in R are described in the [Importing vector graphics](https://cran.r-project.org/web/packages/grImport/vignettes/import.pdf) vignette: *Murrell, P. (2009). Importing vector graphics: The grImport package for R. Journal of Statistical Software, 30(4), 1-37.*

**In addition to installing package `grImport` with the usual `install.packages("grImport")`, [Ghostscript](https://www.ghostscript.com/download/) needs to be installed as well.**

<div id='ReadPS'/>
### Read in the PostScript file

Note: to avoid the ghostscript error 'status 127', the path to ghostscript executable file was given as suggested [here](https://stackoverflow.com/questions/35256501/windows-rgrimport-ghostscript-error-status-127#42393056)

```{r echo = TRUE, message = FALSE}
require(grImport)
Sys.setenv(R_GSCMD = normalizePath("C:/Program Files/gs/gs9.22/bin/gswin64c.exe"))
# Path to example PostScript file
path <- system.file("extdata", "graph_ps.ps", 
                    package = "plotbiomes", 
                    mustWork = TRUE)
# Path to output xml RGML file
RGML_xml_path <- gsub(pattern = "graph_ps.ps", 
                 replacement = "graph_ps.xml", 
                 x = path, 
                 fixed = TRUE)
# Converts a PostScript file into an RGML file
PostScriptTrace(file = path,
                outfilename = RGML_xml_path,
                setflat = 1)
# setflat = 1 assures a visual smooth effect (feel free to experiment)

# Reads in the RGML file and creates a "Picture" object
my_rgml <- readPicture(rgmlFile = RGML_xml_path)
```

<div id='picture'/>
## Visualize and clean (subset) the "Picture" object

```{r echo = TRUE, fig.show = 'hold', out.width = '45%', fig.align = 'center', fig.cap = 'The picture object (left) and its component paths (right). The picture was flipped in Inkscape.'}
# Draws the picture
plot.new(); grid.picture(my_rgml)
# Draws each path composing the picture. This helps selecting desired paths.
plot.new(); picturePaths(my_rgml)
```

By using `picturePaths` one can identify the desired biome lines and the filled polygons. From polygons one can get the fill colors for further use.
```{r echo = TRUE, fig.show = 'hold', out.width = '45%', fig.align = 'center', fig.cap = 'Polygons (left) and their borders (right)'}
# Selects desired paths from the picture object. 
# These are the path corresponding to the filled polygons.
my_fills <- my_rgml[c(3,9,14,16,18,21,23,25,27)]
# Gets the colors
colors <- vector(mode = "character", length = length(my_fills@paths))
for (i in 1:length(my_fills@paths)){
  colors[i] <- my_fills@paths[[i]]@rgb
}

# These are the path corresponding to the biome lines (polygon boundaries)
my_rgml_clean <- my_rgml[c(4,10,15,17,19,22,24,26,28)]
# Assigns colors to the lines
for (i in 1:length(my_rgml_clean@paths)){
  my_rgml_clean@paths[[i]]@rgb <- colors[i]
}

plot.new(); grid.picture(my_fills)
plot.new(); grid.picture(my_rgml_clean)
```

<div id='PScoordinates'/>
## Get the PostScript coordinates from the cleaned "Picture" object

```{r echo = TRUE, out.width = '45%', fig.align = 'center', fig.cap = 'Plot using the original PostScript coordinates'}
x_lst <- vector(mode = "list", length = length(my_rgml_clean@paths))
y_lst <- vector(mode = "list", length = length(my_rgml_clean@paths))
for (i in 1:length(my_rgml_clean@paths)){
  x_lst[[i]] <- my_rgml_clean@paths[[i]]@x
  names(x_lst[[i]]) <- rep(i, length(x_lst[[i]]))
  y_lst[[i]] <- my_rgml_clean@paths[[i]]@y
  names(y_lst[[i]]) <- rep(i, length(x_lst[[i]]))
}
x <- unlist(x_lst)
y <- unlist(y_lst)

par(mar = c(4, 4, 1, 1))
plot(x,y)
```

<div id='conversion_1'/>
## Convert PostScript coordinates into meaningful coordinates

This varies from graph to graph, but in this case luckily there is a grid and axis on the graph that can be used to transform from PostScript coordinates to meaningful coordinates (temperature and precipitations).
```{r echo = TRUE, fig.show = 'hold', out.width = '45%', fig.align = 'center', fig.cap = 'Original grid and axis (left) and their PostScript coordinates (right)'}
my_grid <- my_rgml[1]
my_axis <- my_rgml[30]

x_grid <- my_grid@paths[[1]]@x
y_grid <- my_grid@paths[[1]]@y
x_grid_unq <- unique(sort(x_grid))
y_grid_unq <- unique(sort(y_grid))

# Plot original grid and axis
plot.new()
grid.picture(my_grid)
grid.picture(my_axis)

# Plot in PostScript coordinates
par(mar = c(4, 4, 1, 1))
plot(x_grid, y_grid, pch = 16, col = "red")
points(x, y, cex = 0.1)
abline(v = x_grid_unq[2], col = "red") # X min ~ -10
abline(v = max(my_axis@summary@xscale), col = "red") # X max ~ 30
abline(h = min(my_axis@summary@yscale), col = "red") # Y min ~ 0
abline(h = y_grid_unq[6], col = "red") # Y max ~ 400
```

<div id='conversion_2'/>
### Convert between coordinates systems

Need to identify the corresponding min & max values on OX and OY axis on both coordinates systems.
The conversion equations are those exemplified [here](https://gamedev.stackexchange.com/questions/32555/how-do-i-convert-between-two-different-2d-coordinate-systems).
```{r echo = TRUE, fig.show = 'hold', out.width = '45%', fig.align = 'center', fig.cap = 'Left - original PostScript coordinates; Right - converted coordinates (precipitation vs. temperature)'}
# Notations:
# scs - source coordinate system
# rcs - result coordinate system

x_min_scs <- x_grid_unq[2] # is the X of the left most vertical grid line
x_min_rcs <- -10 # corresponds to -10
x_max_scs <- max(my_axis@summary@xscale) # is the X max of OX axis line
x_max_rcs <- 30 # corresponds to 30

y_min_scs <- min(my_axis@summary@yscale) # is the Y min of OY axis line
y_min_rcs <- 0 # corresponds to 0
y_max_scs <- y_grid_unq[6] # is the Y of the upper most horizontal grid line
y_max_rcs <- 400

# Apply conversion
xx <- (x - x_min_scs)/(x_max_scs - x_min_scs) * (x_max_rcs - x_min_rcs) + x_min_rcs
yy <- (y - y_min_scs)/(y_max_scs - y_min_scs) * (y_max_rcs - y_min_rcs) + y_min_rcs

# There should not be negative values on OY.
# This happens because the original PDF has some artefacts,
# including also some overlapping polygons sections.
min(yy) 
yy[yy < 0] <- 0

# Plot in original coordinates
par(mar = c(4, 4, 1, 1))
plot(x_grid, y_grid, pch = 16, col = "red", 
     xlab = "X - original coordinates",
     ylab = "Y - original coordinates")
points(x, y, cex = 0.1)
abline(v = x_grid_unq[2], col = "red") # X min ~ -10
abline(v = max(my_axis@summary@xscale), col = "red") # X max ~ 30
abline(h = min(my_axis@summary@yscale), col = "red") # Y min ~ 0
abline(h = y_grid_unq[6], col = "red") # Y max ~ 400

# Plot in converted coordinates
plot(xx, yy, cex = 0.1,
     xlab = "X - converted coordinates",
     ylab = "Y - converted coordinates")
abline(v = -10, col = "red") # X min
abline(v = 30, col = "red") # X max
abline(h = 0, col = "red") # Y min
abline(h = 400, col = "red") # Y max
```

<div id='prepare'/>
## Prepare Whittaker_biomes data frame

```{r echo = TRUE}
Whittaker_biomes <- data.frame(temp_c = xx, 
                               precp_cm = yy, 
                               biome_id = as.numeric(names(xx)))

# Assigns the biome names
for (i in 1:nrow(Whittaker_biomes)){
  Whittaker_biomes$biome[i] <- switch(as.character(Whittaker_biomes$biome_id[i]),
                                      "1" = "Tropical seasonal forest/savanna",
                                      "2" = "Subtropical desert",
                                      "3" = "Temperate rain forest",
                                      "4" = "Tropical rain forest",
                                      "5" = "Woodland/shrubland",
                                      "6" = "Tundra",
                                      "7" = "Boreal forest",
                                      "8" = "Temperate grassland/desert",
                                      "9" = "Temperate seasonal forest")
}
```

<div id='colors'/>
### Get the colors also

The fill colors where already extracted from PostScript. In the help of `ggplot2::scale_fill_manual` it is recommended to use a named vector. Therefore, colors were stored in the named character vector `Ricklefs_colors` below:
```{r echo = TRUE}
colors # hexadecimal color codes extracted from PostScript
Ricklefs_colors <- colors
names(Ricklefs_colors) <- unique(Whittaker_biomes$biome)
# Sets a desired order for the names. This will affect the order in legend.
biomes_order <- c("Tundra",
                  "Boreal forest",
                  "Temperate seasonal forest",
                  "Temperate rain forest",
                  "Tropical rain forest",
                  "Tropical seasonal forest/savanna",
                  "Subtropical desert",
                  "Temperate grassland/desert",
                  "Woodland/shrubland")
Ricklefs_colors <- Ricklefs_colors[order(factor(names(Ricklefs_colors), levels = biomes_order))]
Ricklefs_colors
```

<div id='examples'/>
## Examples

Check examples in [Whittaker_biomes_examples](https://rawgit.com/valentinitnelav/plotbiomes/master/inst/doc/Whittaker_biomes_examples.html) vignette, or after package installation, run `RShowDoc("Whittaker_biomes_examples", package = "plotbiomes")`.
