---
title: "Whittaker biomes - check outliers"
author: "Valentin Stefan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    fig_caption: true
---

**In this vignette is discussed about outliers in Whittaker biomes plots using the [`plotbiomes`](https://github.com/valentinitnelav/plotbiomes) package.**

# About outliers

Some precipitation-temperature points may end up outside of the Whittaker biome polygons. This is not unusual and Ricklefs (2008) mentiones that

> "Most of the points fall within a triangular region that includes almost the full range of climates. Only the climates of high mountains do not fall within the triangle." 

(*Ricklefs, R. E. (2008), The economy of nature. W. H. Freeman and Company.*; Chapter 5, Biological Communities, The biome concept; Caption of Figure 5.5)

Nevertheless, one might want to check such points. For example, temperature and precipitation values could be biased because the spatial location was not correctly specified (errors in coordinates). In such cases is very helpful to know the indices of the outliers. This is where the `get_outliers()` and `map_outliers()` functions  from `plotbiomes` package come in handy.

As presented [here][add_pts] one can extract temperature and precipitation values and construct a plot like the one below (note the outliers):

[add_pts]: https://rawgit.com/valentinitnelav/plotbiomes/master/html/Whittaker_biomes_examples.html#add_data_points

```{r, echo = FALSE, message = FALSE}
library(raster)
library(maptools)
# Read temperature and precipitation as raster stack.
# Low resolution raster datasets come with 'plotbiomes' package.
path <- system.file("extdata", "temp_pp.tif", package = "plotbiomes")
temp_pp <- raster::stack(path)
names(temp_pp) <- c("temperature", "precipitation")

data(wrld_simpl) # load world polygons from maptools
# Eliminate Antarctica - it is not represented in the raster datasets.
wrld_simpl <- wrld_simpl[wrld_simpl$NAME != "Antarctica", ]
# Create random locations within world's polygons.
set.seed(66) # random number generator
points <- sp::spsample(x = wrld_simpl, n = 50, type = "random")
# Extract temperature and precipitation values from the raster datasets
extractions <- raster::extract(temp_pp, points, df = TRUE)
# Adjust temperature values to normal scale because WorldClim temperature data
# has a scale factor of 10 (integer storage for saving space).
extractions$temperature <- extractions$temperature/10
# Convert precipitation from mm to cm
extractions$precipitation <- extractions$precipitation/10
```

```{r, echo = FALSE, message = FALSE, out.width = '70%', fig.align = 'center', fig.cap = 'Some points fall outside the triangular region'}

library(plotbiomes)
library(ggplot2)

whittaker_base_plot() +
  # add the temperature - precipitation points
  geom_point(data = extractions,
             aes(x = temperature,
                 y = precipitation),
             size   = 3,
             shape  = 21,
             colour = "gray95",
             bg     = "black",
             stroke = 1,
             alpha  = 0.5) +
  theme_bw()
```

# Get the outliers

The function `get_outliers()` does exactly what its name suggests:

[mpw]: https://github.com/r-spatial/mapview

```{r}
my_outliers <- get_outliers(tp = extractions[, 2:3])
my_outliers
```

Just for fun, here is some code to color/underline the outliers on Whittaker biome plot:

```{r, echo = TRUE, message = FALSE, out.width = '70%', fig.align = 'center', fig.cap = 'Colored outliers'}

library(plotbiomes)
library(ggplot2)

# Color the outliers
extractions$status <- ifelse(extractions$ID %in% my_outliers$row_idx, "out", "in")

# Note that `fill` cannot be used again in the aesthetics 
# because is already in use for biome polygons color filling.
# So one could call `geom_point` two times:
# - once for all points
# - then for placing the colored outliers on top.
whittaker_base_plot() +
  # add the temperature - precipitation points
  geom_point(data = extractions,
             aes(x = temperature,
                 y = precipitation),
             size   = 3,
             shape  = 21,
             colour = "gray95",
             bg     = "black",
             stroke = 1,
             alpha  = 0.5) +
  geom_point(data = extractions[extractions$status == "out",],
             aes(x = temperature,
                 y = precipitation),
             size   = 3,
             shape  = 21,
             colour = "gray95",
             bg     = "red",
             stroke = 1,
             alpha  = 0.5) +
  theme_bw()
```

# Map the outliers

The function `map_outliers()` does exactly what its name suggests. This function makes use of the [mapview][mpw] package functionality to produce interactive views of temperature-precipitation outliers.

```{r echo = FALSE, message = FALSE}
# With RMarkdown, `mapview` needs to be attached to actually make the map appear.
# This should not be needed when scripting though because
# `plotbiomes` package loads `mapview` and there is no need to call `library(mapview)`
library(mapview) 
```

```{r}
map_outliers(tp =  extractions[, 2:3], xy = points@coords)
```

.

`map_outliers()` is a simple wrapper of `mapview::mapView()` function to which arguments can be passed:

```{r}
map_outliers(tp =  extractions[, 2:3], xy = points@coords, 
             col.regions = "red", basemaps = c("OpenTopoMap", "Esri.WorldImagery"))
```
